Identifying covariate traits of trance and possession phenomena in the Ethnographic

### 2022-10-29 Readme.Rmd
finished readme

We want to look at three questions. What sort of social factors have been proposed to co-occur with trance and possession phenomena? Using the Ethnographic Atlas and the Standard Cross-Cultural Sample, can we get more confident that these factors do co-occur with such phenomena across cultures? Using phylogenetic comparative methods, can we make claims on the causative structure between the co-occurent factors?

@@

Many people have argued that trance induced by possession is not a random, idiosyncratic pathology but a recurrent psycho-social trait that is at least partly motivated by other social factors.

High overall social complexity and social rigidity -- the resulting power dynamics -- might translate to possession trance practices. Such practices might serve to express or relay perceived social inflexibility and inequality. This is echoed in the medical anthropology literature (@during2011critical,@dein2012religion) and ties in with broader questions of inequality and mental health (@nguyen2003anthropology). Various causative models have been proposed to account for the emergence of possession trance and, specifically, shamanism in a social setting (@winkelman2015shamanism, @singh2018cultural, @wood2018trance).

@@

Several works made the same point or emphasised various aspects of the same relationship. The above table is somewhat reductive. The argument is spelt out in the main text.

@@

The Atlas records the presence of trance and possession phenomena in a given society, using the following variable (EA112):

@@

This is basically a truth table of whether possession beliefs or trance states are present in a given society, with some causal links between the two (if possession happens it leads to trance, if trance happens it's because of possession, etc). We can think of different groupings (trance phenomena encompass levels 1,3,4,5,6,7, possession phenomena encompass 2,3,4,5,6,7, specifically possession trance covers 3,4,6,7 but not 1,2,5,8).

The Atlas lists a set of variables that tie in with the observed covariates in the literature. These are either taken literally from @bourguignon1968, @bourguignon1973religion, and @greenbaum1973 or adapt variables used by @wood2018complexity, @shaara1992, @winkelman1986, and @douglas2004natural.

@@

## 3. How do we look for correlations between trance and possession and their proposed societal covariates?

@@

Following @bourguignon1968, we reduce the complexity of the outcome variable to two levels: whether a society has *possession trance* or not. The predictors are the possible covariates identified above. Some predictors, like `EA032` (Jurisdictional hierarchy of local community) or `EA113` can be construed to follow a scale. These are input as *ordered factors*. Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are *one-hot encoded*. For details, see the Appendix.

@@

Looking back at the table in the previous section, we see that only about 10% of societies have no trance or possession phenomena. From the rest, a majority have possession trance in some form. Following @bourguignon1968 we can reduce this to a binary: 81 societies have possession trance, 70 have other forms of trance and possessior or show neither.

@@

We now take a look at the variables that were most important in the model itself in predicting the presence of possession trance in a given society, with a cutoff of 10/47.

@@

We now look at the direction of these effects (does e.g.\ increased jurisdictional hierarchy go with more or less prevalent possession trance?).

@@

under construction.

slavery. domestic organisation.
binary.
austronesian (eeh http://lir.byuh.edu/index.php/pacific/article/view/141)
afro-asiatic (sub-saharan, greenbaum)
atlantic-congo (hamito-semitic! lewis)

bayestraits

The independent model assumes the two traits evolve independently, e.g. the transition
from 0 → 1 in the first trait is independent of the state of the second trait. 
dependent: the traits are correlated and the rate of change in one
trait is dependent on the state of the other
100 stones and 1000

mcmc exp 10 priors
iterations per stone to estimate the marginal likelihood

@@

lot of forking paths. slavery in atlantic-congo noteworthy. maybe. overall not really.

@@

These results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature. @bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups. @shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls, @wood2018complexity point to the role of moralising high gods. These are all important variables in our model. @greenbaum1973 and @shaara1992 point to sexual division of labour but this is not relevant in our model.

@@

We can predict some types of trance and possession phenomena in the SCCS subset of the EA. Not all of them. Some of the important co-variates are the ones mentioned by earlier work and the direction of the relationships corroborates the literature.

something something phylogenetic comparative methods

Human society is complicated and cross-cultural data have very low resolution. In this sense, we'd be very suspect of a model that would be too accurate of predicting possession trance from societal covariates. It cannot replace fieldwork and more complex explanatory models either. What it can do is provide an extra leg for such explanatory work, which is broadly the recognised ambition of cross-cultural comparative work in evolutionary anthropology.

@@

>>>>

This is the online supplement for the paper "Possession trance co-varies with measures of social rigidity in the Ethnographic Atlas 2022". The current version of the paper is [here](https://docs.google.com/document/d/1qbcsCeQ-Tw2W-yhQvvYd1Nk5ImQznqWBmiHX09lB_24/edit?usp=sharing).

We have three questions. What sort of social factors have been proposed to co-occur with trance and possession phenomena? Can we give support to these using the Ethnographic Atlas and the Standard Cross-Cultural Sample? Using phylogenetic comparative methods, can we make claims on the causative structure between the co-occurrent factors?

Our informal priors are that trance and possession phenomena result from complex interchanges between indidvidual psychopathology and social organisation, and that, even if we can recognise broader patterns, the effects of social rigidity and marginalisation will be idiosyncratic on the ground. This means that both the outcome (trance and possession) and the predictors (e.g. measures of social rigidity) will be much more loosely defined than in similar analyses. So we expect any effects we find to be relatively modest.

@@

The paper says that trance induced by possession is not a random, idiosyncratic pathology but a recurrent psycho-social trait that is at least partly motivated by other social factors. This is recurrent claim in the literature. Here, we revisit the main points of this argument.

High overall social complexity and social rigidity, as well as the resulting power dynamics, might translate to possession trance practices. Such practices might serve to express or relay perceived social inflexibility and inequality. This is echoed in the medical anthropology literature (@during2011critical,@dein2012religion) and ties in with broader questions of inequality and mental health (@nguyen2003anthropology). Various causative models have been proposed to account for the emergence of possession trance, shamanism, and witchcraft in a social setting (@winkelman2015shamanism, @singh2018cultural, @wood2018trance, @singh2021magic).

@@

Several works made the same point or emphasised various aspects of the same relationship. The above table is somewhat reductive.

@@

What is our outcome variable? The Atlas records the presence of trance and possession phenomena in a given society, using the following variable (EA112):

@@

This is basically a truth table of whether possession beliefs or trance states are present in a given society, with some causal links between the two (if possession happens it leads to trance, if trance happens it is because of possession, etc.). We can think of different groupings: trance phenomena encompass levels 1,3,4,5,6,7, possession phenomena encompass 2,3,4,5,6,7, specifically possession trance covers 3,4,6,7 but not 1,2,5,8. Following @bourguignon1968, we reduce the complexity of the outcome variable to two levels: whether a society has *possession trance* or not.

What are our predictor variables? The Atlas lists a set of variables that tie in with the observed covariates in the literature. These are either taken literally from @bourguignon1968, @bourguignon1973religion, and @greenbaum1973 or adapt variables used by @wood2018complexity, @shaara1992, @winkelman1986, and @douglas2004natural.

@@

## 3. How do we look for correlations between possession trance and its proposed societal covariates?

@@

The predictors are the possible covariates identified above. Some predictors, like `EA032` (Jurisdictional hierarchy of local community) or `EA113` can be construed to follow a scale. These are input as *ordered factors*. Others, like `EA053` (Sex differences in animal husbandry) or `EA043` (Major type of descent) can be carved up into categories: These are *one-hot encoded*. For details, see the Appendix.

@@

Looking back at the table in the previous section, we see that only about 10% of societies have no trance or possession phenomena. From the rest, a majority have possession trance in some form. Following @bourguignon1968 we can reduce this to a binary: 81 societies have possession trance, 70 have other forms of trance and possession or show neither.

@@

We now take a look at the variables that were important in the model itself in predicting the presence of possession trance in a given society, with a cutoff of 10/47 most important variables.

@@

Before we cross-reference the results with the claims in the literature, we look at the direction of these effects (does e.g.\ increased jurisdictional hierarchy go with more or less prevalent possession trance?). Then we ask questions about co-occurrence versus co-evolution.

@@

The literature makes causative claims: societal rigidity causes possession trance (and not the other way round). Cross-cultural correlations can be indicative of causative structure but cannot provide direct evidence for it.

In addition, we can look at the cultural evolution of the outcome and some of the predictor variables. We take two examples, slavery and domestic organisation. If the practice of slavery leads to possession trance, we expect them to (i) correlate in the ethnographic record (they do) and emerge together in cultural phylogenies.

To test for this, we select two predictors, slavery and domestic organisation. These are the two strongest predictors in our cross-cultural analysis. We recode them as binaries to reflect their relationship to prevalence of possession trance: For prevalence of slavery, 1 is any form of slavery present in a cultural group, 0 is no slavery attested. For domestic organisation, 1 is nuclear family organisation, 0 is larger extended family organisation. Our outcome is 1 where possession trance is present and 0 where it is absent.

We select three phylogenies, Austronesian, Afro-Asiatic, and Atlantic-Congo. These are robust phylogenies with a large number of societies in each. They include societies that have been discussed by the trance and possesion literature (@blust2000rat, @greenbaum1973, @boddy1989wombs, @lewis1991introduction).

We use BayesTraits to assess the co-evolution of predictors and outcome (@pagel2007bayestraits). We fit two models, an independent model, which assumes that two traits evolve independently, and a dependent model, which assumes that traits are correlated and the rate of change in one trait. is dependent on the state of the other trait.

Six dependent and six independent models are fit, for the relationship of (i) possession trance and the presence of slavery, (ii) possession trance and nuclear/extended family organisation across the three phylogenies.

Phylogenies come from D-Place. Societies are matched to phylogenetic trees by language ID-s. One language is used by more than one society in Afro-Asiatic and in Atlantic-Congo respectively, these are removed. Societies that have no data on both traits are pruned from the phylogenetic trees.

Models are fit using Markov Chain Monte Carlo using exp 10 priors, 100 stepping stones, and 1000 iterations per stone. Each run provides a marginal likelihood for the model fit.

We calculate the Bayes Factors for the more complex dependent versus the less complex independent models. A higher value for a Bayes Factor indicates that the more complex model provides a better fit and so its complexity is justified. The Bayes Factors for the six model comparisons can be seen below.

@@

The results strongly suggest that possession trance and slavery as well as nuclear family organisation developed independently across the three phylogenies. One apparent exception is slavery and possession trance in Afro-Asiatic languages, which of course cover the famed description of possession trance practices in the Sudan and the Horn of Africa. But a Bayes Factor of 4 is not substantial. Overall, there is little evidence here for trait co-evolution.

The phylogenetic comparative analysis concedes with a large number of limitations. It remains true that it does not lend additional support to the cross-cultural analysis.

@@

The overall results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature. @bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups. @shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls, @wood2018complexity point to the role of moralising high gods. These are all important variables in our cross-cultural model. @greenbaum1973 and @shaara1992 point to sexual division of labour but this is not relevant in our model.

@@

We can predict some types of trance and possession phenomena in the Standard Cross-Cultural Sample subset of the Ethnographic Atlas. Not all of them. Some of the important co-variates are the ones mentioned by earlier work and the direction of the relationships corroborates the literature. Phylogenetic comparative methods do not provide additional support to attested cross-cultural co-variation.

Human society is complicated and cross-cultural data have very low resolution. In this sense, we'd be suspect of a model that would be too accurate of predicting possession trance from societal covariates. It cannot replace fieldwork and more complex explanatory models. It can provide an extra leg for such explanatory work, which is the recognised ambition of cross-cultural comparative work in evolutionary anthropology.

@@

## 9. Limitations

The cross-cultural model is fit on the Standard Cross-Cultural Sample which does not minimise the phylogenetic signal completely (see [here](https://excd.org/2018/03/01/stats-corner-is-the-standard-cross-cultural-sample-really-standard/)). The data in the Ethnographic Atlas cannot be independently verified. The outcome variable and the predictors are selected and defined with the existing literature in mind. It remains true that these decisions use up a large amount of researcher degrees of freedom. This is especially true for the phylogenetic comparative analysis, which further reduces resolution on the co-varying traits as a starting point.

Such limitations will be in place for all cross-cultural analyses. They are exacerbated here by the elusiveness of the variables themselves, which likely results in more starting variability baked into the Ethnographic Atlas coding in the first place. The accompanying statistical rigour might render this work relevant to the broader study of trance and possession, despite these limitations.
